---
title: "Snowshoe Hares - Exploratory Report"
author: "Kirsten White"
date: "November 9, 2021"
output: 
  html_document: 
    theme: flatly
    toc: yes
    highlight: textmate
    code_folding: hide
---

# **Introduction** 

This report provides an exploratory overview of juvenile snowshoe hare populations observed in the Bonanza Creek Experimental Forest in Alaskan boreal forests from 1998 to 2012 (Kielland, K, et al 2017). Data was collected on showshoe hares (Lepus americanus) by Knut Kielland and colleagues in Bonanza Creek Experimental Forest, part of the US Long Term Ecological Research Network (LTER). Capture-recapture of snowshoe hares was conducted at 5 locations in the Tanana Valley in an effort to evaluate population fluctuations within the keystone prey species. In this report, total juvenile trap counts are evaluated by year and particular characteristics are evaluated. More specifically, male and female weight is compared and the relationship between hare weight and hind foot length is explored. 

# **Data & Analysis** 

The data in this report contains observations for 371 juvenile hares gathered between 1998 and 2012. Following exploratory data visualization, juvenile male and female hare weights are evaluated with a two-sample t-test using a significance level of 0.05 throughout.  Differences are explored using Cohen's d effect size nd the relationship between weight and hindfoot length is investigated using simple linear regression. All analyses are in R Version 4.1.1 using R studio version 1.4.1717. 

**Data Citation**: Kielland, K., F.S. Chapin, R.W. Ruess, and Bonanza Creek LTER. 2017. Snowshoe hare physical data in Bonanza Creek Experimental Forest: 1999-Present ver 22. Environmental Data Initiative. https://doi.org/10.6073/pasta/03dce4856d79b91557d8e6ce2cbcdc14


```{r setup, message = FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

### Setup

# --- Attach Packages ---
library(tidyverse)
library(here)
library(janitor)
library(lubridate) 
library(kableExtra)
library(effsize)
library(beeswarm)
```


```{r}
### Read in the Data
bonanza_hares <- read_csv(here("data", "bonanza_hares.csv"))
```
---
# **Results**

## **Part 1: Juvenile Hare Trapping Counts**
Total number of juvenile snowshoe hares trapped during each year of the study (1998 - 2012) are visualized in a bar graph and a summary stastitics table.  

```{r}
### Data Wrangling 
# Juvenile hare trappings are counted by year. 

# A dataset including only juvenile hares, a new year column, and counting total juveniles per year is created 
juvenile_trappings <- bonanza_hares %>%
  mutate(date = mdy(date)) %>%
  mutate(year = year(date)) %>%
  filter(age == "j") %>%
  group_by(year, age) %>%
  count(age)
```



```{r}
### Data Visualization 
# Juvenile hare trappings by year are visualized in a bar graph.

# A bar graph is plotted with counts on the y-axis and year on the x-axis
ggplot(data = juvenile_trappings, aes(x = year, y = n)) +
  geom_bar(stat = "identity", fill = "aquamarine4", color = "grey", width = 0.5) + 
# Graph title, axis, and caption are labeled 
  scale_y_continuous(breaks = c(0, 25, 50, 75, 100, 125))+
  scale_x_continuous(breaks = c(1998, 2000, 2002, 2004, 2006, 2008, 2010, 2012))+
  labs(x = "Year", 
       y = "Hare Count", 
       title = "Total Juvenile Snowshoe Hares Trapped by Year (1998 - 2012) (WHITE)",
       caption = "Data Source: Kielland, K., F.S. Chapin, R.W. Ruess, and Bonanza Creek LTER. 2017") + 
# Title, axis, and caption font, size, and placement are edited 
  theme_grey() + 
  theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 12)) + 
  theme(axis.title = element_text(face = "bold")) +
  theme(plot.caption = element_text(hjust = 1, size = 8))+
  theme_update()
```

**Figure 1**: Total number of juvenile hares trapped per year from 1998 through 2012. No counts were collected in 2002 and 2009.



**Table 1**:  Summary statistics for juvenile hare counts from 1998 - 2012


```{r}
### Summary Table & Discussion
# Minimum, maximum, mean, median, and standard deviation of juvenile hare counts are calculated and visualized in a table. 

# Create a summary statistic table of juvenile hare counts by year
juvenile_trappings_summary <- juvenile_trappings %>%
  group_by(age) %>%
  summarize(sample_size = n(), 
            min(juvenile_trappings$n, na.rm = TRUE), 
            max(juvenile_trappings$n, na.rm = TRUE), 
            mean(juvenile_trappings$n, na.rm = TRUE), 
            median(juvenile_trappings$n, na.rm = TRUE), 
            sd(juvenile_trappings$n, na.rm = TRUE)) %>%
  kbl(col.names = c("Age", "Sample Size", "Min", "Max", "Mean", "Median", "Standard Deviation"), digits = 1) %>%
  kable_classic_2()

juvenile_trappings_summary
```

**Analysis**:  Over the 12 year period, total number of trappings varied greatly. At the beginning of the study (1998-1999), significantly more juveniles were trapped reaching a maximum of 126 hares per year and steadily decreased over time to a minimum of 2 hares trapped per year. The mean number of juveniles trapped per year was 31.5 while the median value, much more representative, was 18.5 hares trapped per year; total standard deviation was 35.9 hares trapped per year. See **Figure 1**, **Table 1**. In future analysis, data can be standardized by having equal sample sizes.

---

##  **Part 2: Juvenile Hare Weights**
Juvenile snowshoe hare weights are compared by sex and site and visualized in a scatterplot.  

```{r}
### Data Wrangling 
# Data is wrangled to include observations on juvenile hares only and information on sex, weight, and grid (site). 

# Create a new data set including only columns for sex, weight, and site (grid) and including observations of only juvenile hares
juvenile_weights <- bonanza_hares %>%
  filter(age == "j") %>%
  drop_na(weight) %>%
  select(sex, weight, grid) %>%
# Add a new column with the full site name 
  mutate(grid_full = case_when(
        grid == "bonbs" ~ "Bonbs", 
        grid == "bonmat" ~ "Bonanza Mature", 
        grid == "bonrip" ~ "Bonanza Riparian")) %>%
# Update the sex column to capitalized M and F for Male and Female 
  mutate(sex = case_when(
        sex == "m" ~ "M", 
        sex == "f" ~ "F"))
```


```{r}
### Data Visualization 
# Juvenile male and female weights are visualized by site. 

# Create a scatterplot showing male and female juvenile hare weights separated by site
ggplot(data = juvenile_weights, aes(x = sex, y = weight, color = grid_full))+
  geom_jitter(width = 0.1, show.legend = "none")+
  facet_wrap(~ grid_full) +
  scale_color_manual(values = c("aquamarine4", "goldenrod2", "tomato4"))+
   labs(x = "Sex", 
       y = "Weight (g)", 
       color = "Site",
       title = "Juvenile Snowshoe Hare Weight by Sex and Site (WHITE)", 
       caption = "Data Source: Kielland, K., F.S. Chapin, R.W. Ruess, and Bonanza Creek LTER. 2017") + 
  theme(plot.title = element_text(face = "bold"))+
  theme(plot.caption = element_text(hjust = 1, size = 8))+
  theme(axis.title = element_text(face = "bold")) 
```


**Figure 2**: Female and male juvenile hare weights are visualized by sex (female, male, and NA) and site, including: Mature (green), Riparian (gold), and Bonbs (maroon). Each point indicates an individual observation for a juvenile hare weight.

**Table 2**: Summary statistics for male and female juvenile weights (g) separated by site
```{r}
### Summary Table & Discussion
# Mean, median, minimum, maximum, and standard deviation of male and female juvenile hare weights are calculated and visualized in a table.

juvenile_weights_summary <- juvenile_weights %>%
  group_by(grid_full, sex) %>%
  drop_na(weight) %>%
  summarize(sample_size = n(), 
            mean_weight = mean(weight, na.rm = TRUE), 
            median_weight = median(weight, na.rm = TRUE), 
            min_weight = min(weight, na.rm = TRUE), 
            max_weight = max(weight, na.rm = TRUE),
            sd_weight = sd(weight, na.rm = TRUE)) %>%
  kbl(col.names = c("Site", "Sex","Sample Size", "Mean (g)", "Median (g)", "Min (g)", "Max (g)", "Standard Deviation (g)"), digits = 1) %>%
  kable_classic_2()

juvenile_weights_summary
```

**Analysis**: While hare weights recorded at the Bonbs site appear to have higher weights (both in mean and median), it is important to note the large disparity in sample sizes is evident between each site.  Counts from the Bonanza Riparian site had a significantly higher sample size (242 hares) compared to the other two sites (82 and 47 hares). Additionally, spread in weight value is high between each site and both sexes. 

```{r, include=FALSE}
### Statistical Analysis

# Run a T-test,  Mann Whitney U unpaired rank-based test, and Cohen's D

# Create vector of female weights 
female_weights <- juvenile_weights %>%
  filter(sex == "F") %>%
  drop_na(weight) %>%
  pull(weight)

# Create vector of male weights
male_weights <- juvenile_weights %>%
  filter(sex == "M") %>%
  drop_na(weight) %>%
  pull(weight)

# Plot histogram and ggplots to check assumptions
hist(female_weights)
hist(male_weights)
qqnorm(female_weights)
qqnorm(male_weights)

# Find means and standard deviation for male and female weights

male_mean <- mean(male_weights)
female_mean <- mean(female_weights)
male_sd <- sd(male_weights)
female_sd <- sd(female_weights)

# Run a two sample t-test on female versus male weights 
# Asking the question: What is the probability that I would have found samples with means that are *at least as different as the ones I've found* by random chance, if they are drawn from populations with the same mean?
weight_t_test <- t.test(female_weights, male_weights)
weight_t_test # p = 0.7%

# Run a Mann Whitney U unpaired rank-based test to compare ranks (medians) between two unpaired samples
# This test investigates the difference in ranks (also called median comparisons) to determine if there is a signficant difference in ranks between the weight values of male and female juvenile hares. 
weight_mwu <- wilcox.test(female_weights, male_weights)
weight_mwu # p = 0.7%

# Run a Cohen test
weight_cohen <- cohen.d(female_weights, male_weights)
weight_cohen #

```
---
## **Part 3: Juvenile Weight Comparison**
**Table 2**: Summary statistics for male and female juvenile weights (g) by site, including: sample size, mean, median, min, max, and standard deviation. 

```{r}
# Compare juvenile weights in males and females

juvenile_weights_m_f <- juvenile_weights %>%
  group_by(sex) %>%
  drop_na(weight) %>%
  summarize(sample_size = n(), 
            mean_weight = mean(weight, na.rm = TRUE), 
            median_weight = median(weight, na.rm = TRUE), 
            min_weight = min(weight, na.rm = TRUE), 
            max_weight = max(weight, na.rm = TRUE),
            sd_weight = sd(weight, na.rm = TRUE)) %>%
  kbl(col.names = c("Sex","Sample Size", "Mean (g)", "Median (g)", "Min (g)", "Max (g)", "Standard Deviation (g)"), digits = 1) %>%
  kable_classic_2()

juvenile_weights_m_f
```

**Analysis**: On average, male juvenile hares weigh more than female juvenile hares. The actual difference in mean weight is `r round(male_mean - female_mean, 1)` (g) and the percent difference is `r round((male_mean - female_mean)/((male_mean + female_mean)/2) * 100, 1)`%. In addition, the difference in means was found to be statistically significant (Welch’s two-sample t-test:  t(`r round(weight_t_test$parameter,1)`) = `r round(weight_t_test$statistic, 1)`,  p < 0.05) with a small effect size (Cohen’s d = `r round(weight_cohen$estimate, 2)`). Therefore, further investigation is needed to identify whether male snowshoe hares have an actual significant difference in weight compared to female snowshoe hares. 


```{r, include = FALSE}
### Exploratory graphs

ggplot(data = juvenile_weights, aes(x = weight)) +
  geom_histogram(bins = 10) +
  facet_wrap(~ sex)

ggplot(data = juvenile_weights, aes(sample = weight)) +
  geom_qq() +
  facet_wrap(~ sex)
```

---

## **Part 4: Juvenile Hare Weight & Hind Foot Length**
The relationship between juvenile snowshoe hare hind foot length and weight is explored, visualized, and analyzed. 


```{r}
### Data Wrangling
# Data is wrangled to include only observations for juvenile hares with information on their hind foot length and weight. 

# Weight and hindfoot data is selected. Data is filtered to include Observations for juveniles only. 
juvenile_hfoot <- bonanza_hares %>%
  filter(age == "j") %>%
  select(weight, hindft)
```


```{r}
### Data Visualization 
# Hindfoot versus weight data for juvenile hares is visualized. 

ggplot(data = juvenile_hfoot, aes(x = hindft, y = weight)) +
  geom_jitter() +
  geom_smooth(method = "lm", color = "aquamarine4", fill = "goldenrod1") +
  labs(x = "Hindfoot Length (mm)", 
       y = "Weight (g)", 
       title = "Juvenile Hare Weight and Hindfoot Comparison (WHITE)", 
       caption = "Data Source: Kielland, K., F.S. Chapin, R.W. Ruess, and Bonanza Creek LTER. 2017") +
  theme(plot.title = element_text(hjust = 0.4, face = "bold"))+
  theme(plot.caption = element_text(hjust = 1, size = 8))+
  theme(axis.title = element_text(face = "bold"))
```


**Figure 3**: A simple linear regression was used to explore the relationship between hind foot length (mm) and body weight (g) across all juvenile snowshoe hares (male and female).
 
**Linear Model Analysis**: Based on this initial impression, the relationship appears heteroscedastic since the spread of residuals about the model changes in a non-random way (more outliers) as hind foot length increases and weight becomes less predictable.

```{r, include = FALSE}
# Linear model, stored as hfoot_weight_lm
hfoot_weight_lm <- lm(weight ~ hindft, data = juvenile_hfoot)

# Return complete overview
summary(hfoot_weight_lm)

# Use broom::tidy() function to see the model outputs in a nice format
hfoot_weight_lm_tidy <- broom::tidy(hfoot_weight_lm)

hfoot_weight_lm_out <- broom::glance(hfoot_weight_lm)
hfoot_weight_lm_out

# Find Pearson's r value which indicates the strength of the correlation (degree of relationship) between weight and hindfoot length 
hfoot_weight_cor <- cor.test(juvenile_hfoot$weight, juvenile_hfoot$hindft)

```

- The **slope** is `r round(hfoot_weight_lm$coefficient[2], 2)` (g/mm) and the **y-intercept** is `r round(hfoot_weight_lm$coefficient[1], 1)`  (g). The full equation is mass = `r round(hfoot_weight_lm$coefficient[2], 2)`*(hindfoot length) + (`r round(hfoot_weight_lm$coefficient[1], 1)`). 

- The **R^2^ value** of `r round(hfoot_weight_lm_out$r.squared, 1)` means that `r round(hfoot_weight_lm_out$r.squared * 100)`% of the variance in hind foot length length is explained by weight. **Pearson's r correlation** is `r round(hfoot_weight_cor$estimate, 1)`. This R^2^ value indicates that hind foot length has a medium to moderate influence on hare weight. 

---

# **Summary** 
Exploratory data analysis reveals the following: 

  - The total count of juvenile hares declined over the time period of this study, 1998 -2012. 
  - Overall, male juvenile hares weigh more than female juvenile hares. 
  - A moderate correlation was detected between weight and hind foot length, but further analysis is required to confirm this. 
  
  
---

# **Citations**
  
1) Data citation: Kielland, K., F.S. Chapin, R.W. Ruess, and Bonanza Creek LTER. 2017. Snowshoe hare physical data in Bonanza Creek Experimental Forest: 1999-Present ver 22. Environmental Data Initiative. https://doi.org/10.6073/pasta/03dce4856d79b91557d8e6ce2cbcdc14
  